Safe Transferable Regions

### 2016-07-03 paper/abstract.tex
abstract and intro
Wibbles plus new reference

Manual memory management is susceptible to the usual perils of
dangling pointers. A key contribution of this paper is a
refinement-based region type system that ensures the memory safety of
C\# programs in the presence of transferable regions. We complement
our type system with a type inference algorithm that infers principal
region types for first-order programs, and practically useful types
for higher-order programs. This eliminates the need for programmers to
write region annotations on types, while facilitating the reuse of
existing C\# libraries with no modifications. Experiments demonstrate
the practical utility of our approach.

>>>>

Manual memory management is susceptible to the usual perils of
dangling pointers. A key contribution of this paper is a
refinement-based region type system that ensures memory safety
in the presence of transferable regions. We complement
our type system with a type inference algorithm that infers principal
region types for first-order programs, and practically useful types
for higher-order programs. This eliminates the need for programmers to
write region annotations on types, while facilitating the reuse of
existing libraries with few modifications. Experiments demonstrate
the expressiveness and practical utility of our approach.

### 2016-07-03 paper/introduction.tex
Wibbles plus new reference

In this paper, we present a \emph{safe} and \emph{efficient} region-based approach
to \emph{data transferable} between a set of concurrent actors, motivated by the following concerns.

Consider the example (of  a producer creating a data structure and transferring it to a consumer)
in Fig.~\ref{fig:motivating-eg}. This represents a \C{SELECT} query operator for streaming data. 
The operator receives a stream of input messages, each
associated with a time (window) $t$, processed by method
\texttt{onReceive}.  Each input message contains a list of inputs,
each of which is processed by applying a user-defined function to
create a corresponding output.  The operator may receive multiple
messages with the same timestamp (and messages with different
timestamps may be delivered out of order). A timing-message (an
invocation of method \texttt{OnNotify}) indicates that no more input
messages with a timestamp $t$ will be subsequently delivered.  At this
point, the operator completes the processing for time window $t$ and
sends a corresponding output message to its successor.

@@

\paragraph{Performance.}
When the transferred data structures are large, as is the case with many streaming big-data
analysis systems, garbage collection overhead becomes
significant~\cite{Broom:HotOS}.  Furthermore, in a distributed
dataflow system, the GC pause at one node can have a cascading adverse
effect on the performance of other nodes~\cite{Broom:HotOS,harris15}:
a GC pause at an upstream actor can block downstream actors that are waiting for
messages.  However, often much of the GC overhead results from the
collector performing avoidable or unproductive work.  For the example
in Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the
\C{map} data structure, although its objects cannot be collected until a
suitable timing message arrives.

\begin{figure}
\centering
\subcaptionbox {
  Before transfer.
  \label{fig:after-transfer}
} [
  0.35\columnwidth
] {
  \includegraphics[scale=0.36]{producer-heap}
}
\subcaptionbox {
  After transfer.
  \label{fig:before-transfer}
} {
  \includegraphics[scale=0.36]{producer-consumer-heap}
}

\caption{References in and out of a transferable region ($R$) become
invalid after transfer.}
\label{fig:transfer-violations}
\end{figure}

\paragraph{Safety.}
These systems are designed to allow the producer and consumer to
execute in the same address space or in different address spaces
(as determined by the compiler and runtime).
We wish to ensure memory safety in the presence of such transferred data by ensuring
that there are no invalid references. A transfer operation, with no
additional checks, may cause memory safety violations, both at the
producer of the data structure, and its (possibly remote) consumer. At
the producer, any existing references into the transferred data structure become
invalid post transfer. If the data structure contains references to
objects outside (the transferred data), then such references become
invalid in the context of the consumer. Both scenarios are depicted in
Fig.~\ref{fig:transfer-violations}. Safety violations of this kind are
particularly unwelcome in the context of languages like C\# and Java
that are otherwise memory-safe. While references from within the
transferable data to outside can be disallowed in the interest of safety,
references into transferable data needs to be allowed before the data
is transferred. Allowing such references is crucial to performance,
as any non-trivial program creates temporary references to the internal
objects of a data-structure.

\paragraph{Safe Transferable Regions.}
In this paper, we present a language-based approach to safely and
efficiently realize such transferable data that uses regions as
the implementation technique.
%
A region is a block of memory that is allocated
and freed in one shot, often in constant time. A region may contain
one or more contiguous range of memory locations, and individual
objects may be dynamically allocated within the region over time,
while they are deallocated en masse when the region is freed.  Thus, a
region is a good fit for a transferable data-structure.  In
Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
time window $t$ (i.e., \C{map[t]}) can be a separate region that is
allocated when the first message with timestamp $t$ arrives, and
deallocated after \C{map[t]} is transferred in \C{onNotify}.

The use of regions alleviates the performance concern related to
garbage collection described earlier. (It also enables more efficient
serialization/deserialization of data-structures across address spaces,
which is also a significant performance bottleneck in such systems.)
However, manual region memory management introduces memory safety
issues such as dangling pointers.
We present a single type system that simultaneously addresses
this safety concern, as well as those described above related to
the use of \emph{transferable data}.

% Region-based memory management, both manual as well as automatic, has
% been known for a long time. Manual region-based memory management
% suffers from the usual drawbacks, namely the potential for invalid
% references and the consequent lack of memory safety. Automatic
% region-based memory management systems guarantee memory safety, but
% impose various restrictions.  MLKit, which implements the approach
Safe region-based memory management using types was
pioneered by Tofte and Talpin~\cite{tofte94,tofte97}, who 
use only lexically scoped regions.  At runtime, the set of all regions
(existing at a point in time) forms a stack. Thus, the lifetimes of
all regions must be well-nested: it is not possible to have two
regions whose lifetimes overlap, with neither one's lifetime contained
within the other.  Unfortunately, the data structures in the above
example do not satisfy this restriction (as the output messages for
multiple time windows may be simultaneously live, without any
containment relation between their lifetimes).  We refer to regions
with lexically scoped lifetimes as \emph{stack regions} and to regions
that do not have such a lexically scoped region as \emph{dynamic}
regions.

Our focus, in this paper, is on first-class \emph{memory-safe} dynamic regions
that can be safely transferred across address spaces. We refer to such dynamic
regions as \emph{transferable} regions.
%
Our approach is based on a variation of ideas introduced by~\cite{WW01,cyclone04}
that combine linear types and regions to suport dynamic regions.
Unlike the prior work, we ensure memory safety in the presence of
transferable regions through a \emph{combination of a static
typing discipline and lightweight runtime checks in place of linear types}.

The cornerstone of this approach is an \C{open} lexical block for transferable regions,
that ``opens'' a transferable region and guarantees that the region
won't be transferred/freed while it is open. By nesting an
MLKit-style \C{letregion} lexical block, that delimits
the lifetime of a stack region, inside an \C{open} lexical block for a
transferable region, we can guarantee that the transferable region
will remain live as long as the stack region is live. We say that the
former \emph{outlives} the later., and any references from the stack
region to the transferable region are therefore safe.  Next, we note
that by controlling the outlives relationships between various
regions, we only allow safe cross-region references, while prohibiting
unsafe ones. In the above example, an outlives relationship
\emph{from} the stack region \emph{to} the transferable region means
that the references in that direction are allowed, but not the
references in the opposite direction. In contrast, if an \C{open}
block of a transferable region $R_0$ is nested inside an \C{open}
block of another transferable region $R_1$, we do not establish any
outlives relationships, thus declaring our intention to not allow any
cross-region references between $R_0$ and $R_1$.  Finally, we observe
that outlives relationships are established based on the lexical
structure of the program, hence a static type system can enforce them
effectively. By assigning region types to objects, which capture the
regions such objects are allocated in, and by maintaining outlives
relationships between various regions, we can statically decide the
safety of all references in the program.

Formally, our type system introduces \emph{region parameters} (for both
classes and methods) and uses constrained parametric polymorphism
over these parameters, where the constraints capture outlives constraints
between the region parameters.

Ensuring memory safety using this approach requires ensuring
that the use of transferable regions satisfies certain temporal properties.
Firstly, a transferable region should not be transferred/freed
inside an open block of that region (i.e, while it is still open).
Secondly, a transferred/freed region should not be opened. These are
typestate invariants on the transferable region objects, which are
hard to enforce statically in the presence of unrestricted
aliasing. Techniques like linear types and unique pointers can be used
to restrict aliasing, but the constraints they impose are often hard
to program around. We therefore enforce typestate invariants at
runtime via lightweight checks. In particular, we define an acceptable
state transition discipline for transferable regions
(Fig.~\ref{fig:region-fsm}), and check, at runtime, whether a given
transition of a transferable region (\eg, from \emph{open} state to
\emph{freed} state) is valid or not. The check is lightweight since it
only involves checking a single tag that captures the current state.
We believe that this is a reasonable choice since regions are
coarse-grained objects manipulated infrequently, when compared to the
fine-grained objects that are present inside these regions, for which
safety is enforced statically.

One of the novel contributions of this paper is a sound type inference algorithm
that eliminates the need for users to write region type annotations.
This, we believe, significantly reduces the impediment to adopt our
approach in practical setting.
Our inference algorithm proceeds in two stages: in the first stage,
we introduce region parameters (for classes and methods), and in
the second stage we infer the outlives constraints between these
parameters (using an interprocedural fixed point computation).
% While the first

@@

% , and it is undesirable to allocate them in a transferable
% region; such regions are meant for the data being transferred.

%As with allocation and deallocation, transferring a memory region is fast, and therefore,
%transferring a data structure contained in a region is more efficient
%that traversing its objects in heap, and transferring them
%independently.
%In the \C{SelectVertex} example from
%Fig.~\ref{fig:motivating-eg}, the proposed region to contain the
%output for each time window $t$ must be transferable due to the
%\C{transfer} operation on Line 16. As it is the case with
%\C{SelectVertex}, the transferred data is no longer accessed by the
%producer, so the transfer operation in our system deallocates the
%region once the transfer is complete.
%
%Since transferable regions are first class objects in our setting,
%controlling references to and from such regions while ensuring their
%safety without significantly diluting the performance advantage of
%regions over GC is a challenging exercise.
%
%An important observation, in the context of processes of the kind
%described above, is that the data-structures exchanged between them
%can be partitioned into sets of fate-sharing objects with common
%lifetimes, which makes them good candidates for a region-based memory
%management discipline.

>>>>

Consider the example, from~\cite{Broom:HotOS}, shown in
Fig.~\ref{fig:motivating-eg}.  This code represents the logic for a
streaming query operator.  The operator receives a stream of input
messages, each associated with a time (window) $t$, processed by
method \texttt{onReceive}.  Each input message contains a list of
inputs, each of which is processed by applying a user-defined function
to create a corresponding output.  The operator may receive multiple
messages with the same timestamp (and messages with different
timestamps may be delivered out of order).  A timing-message (an
invocation of method \texttt{OnNotify}) indicates that no more input
messages with a timestamp $t$ will be subsequently delivered.  At this
point, the operator completes the processing for time window $t$ and
sends a corresponding output message to its successor.

@@

This example is an instance of a general pattern, where a producer creates a
data structure and passes it to a consumer. In a system where most of the
computation takes this form, and these data structures are very large, as is
the case with many streaming big-data analysis systems, garbage collection
overhead becomes significant~\cite{Broom:HotOS,Nguyen+:facade}.  Furthermore, in a
distributed dataflow system, the GC pause at one node can have a cascading
adverse effect on the performance of other nodes, particularly when real-time
streaming performance is required~\cite{Broom:HotOS, harris15}.  In
particular, a GC pause at an upstream actor can block downstream actors that
are waiting for messages.  However, often much of the GC overhead results
from the collector performing avoidable or unproductive work.  For example,
in the process executing the code from Fig.~\ref{fig:motivating-eg}, GC might
repeatedly traverse the \C{map} data structure, although its objects cannot
be collected until a suitable timing message arrives.

% actors.
% performance degradation due to
% GC overhead becomes aggravated in the distributed 
% For example, a consumer process may trigger a
% GC as soon as it receives a large data structure from a producer
% process, which has just finished its GC. In such case, consumer's GC
% traverses the entire data structure performing little or no work.

% as the collector scans these large data structures For instance, in
% the above example, the list at \C{map[t]} cannot be collected until
% the the timing message arrives, and the list is transferred.
% \C{transfer} operation is complete.  using alternative memory
% management techniques.

An important observation, in the context of processes of the kind
described above, is that the data-structures exchanged between them
can be partitioned into sets of fate-sharing objects with common
lifetimes, which makes them good candidates for a region-based memory
management discipline. A region is a block of memory that is allocated
and freed in one shot, consuming constant time. A region may contain
one or more contiguous range of memory locations, and individual
objects may be dynamically allocated within the region over time,
while they are deallocated en masse when the region is freed.  Thus, a
region is a good fit for a set of fate-sharing objects.
% that is used to realize a data structure with a well-defined
% lifetime. 
In Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
time window $t$ (i.e., \C{map[t]}) can be a separate region that is
allocated when the first message with timestamp $t$ arrives, and
deallocated after \C{map[t]} is transferred in \C{onNotify}.

Region-based memory management, both manual as well as automatic, has
been known for a long time. Manual region-based memory management
suffers from the usual drawbacks, namely the potential for invalid
references and the consequent lack of memory safety. Automatic region-based memory
management systems guarantee memory safety, but impose various
restrictions.  MLKit, which implements the approach pioneered by Tofte
and Talpin~\cite{tofte94,tofte97}, for example, uses lexically scoped
regions.  At runtime, the set of all regions (existing at a point in
time) forms a stack. Thus, the lifetimes of all regions must be
well-nested: it is not possible to have two regions whose lifetimes
overlap, with neither one's lifetime contained within the other.
Unfortunately, the data structures in the above example do not satisfy
this restriction (as the output messages for multiple time windows may
be simultaneously live, without any containment relation between their
lifetimes).  We refer to regions with lexically scoped lifetimes as
\emph{stack regions} and to regions that do not have such a lexically
scoped region as \emph{dynamic} regions.

The goal of this work is a \emph{memory-safe} region-based memory
management technique that supports dynamic regions as first-class
objects. Our focus, in this paper, is on dynamic regions, which can be
safely transferred across address spaces. We refer to such dynamic
regions as \emph{transferable} regions. As with allocation and
deallocation, transferring a memory region is fast, and therefore,
transferring a data structure contained in a region is more efficient
that traversing its objects in heap, and transferring them
independently\footnote{Empirical studies in~\cite{Broom:HotOS} support
this claim}. In the \C{SelectVertex} example from
Fig.~\ref{fig:motivating-eg}, the proposed region to contain the
output for each time window $t$ must be transferable due to the
\C{transfer} operation on Line 16. As it is the case with
\C{SelectVertex}, the transferred data is no longer accessed by the
producer, so the transfer operation in our system deallocates the
region once the transfer is complete.

With respect to memory safety, the key property we wish to ensure is
that there are no invalid references: i.e., references to objects that
were deallocated, or simply never existed. Transfer operation, with no
additional checks, may cause memory safety violations, both at the
producer of the data structure, and its (possibly remote) consumer. At
the producer, any existing references into the data structure become
invalid post transfer. If the data structure contains references to
objects outside its (transferable) region, then such references become
invalid in the context of the consumer. Safety violations of this kind
are particularly unwelcome as the program with GC did not have them to
begin with. Note that, while the references of later kind (i.e.,
references that escape a region) defeat the very purpose of a
transferable region, hence need to be prohibited, references of the
former kind (i.e., references into a region from outside) are not at
odds with the concept of a transferable region, hence need to be
permitted. In fact, allowing such references is crucial to
performance, as any non-trivial program creates temporary objects, and
it is undesirable to allocate them in a transferable region; such
regions are meant for the data being transferred. Since transferable
regions are first class objects in our setting, controlling references
to and from such regions while ensuring their safety without
significantly diluting the performance advantage of regions over GC is
a challenging exercise.

In this paper, we describe an approach that restores memory safety in
presence of transferable regions through a combination of a static
typing discipline and lightweight runtime checks. The cornerstone of
our approach is an \C{open} lexical block for transferable regions,
that ``opens'' a transferable region and guarantees that the region
won't be transferred/freed while it is open. Our observation is that
by nesting a~\cite{tofte94}-style \C{letregion} lexical block, that
delimits the lifetime of a stack region, inside an \C{open} lexical
block for a transferable region, we can guarantee that the
transferable region will remain live as long as the stack region is
live. We say that the former ``outlives'' the later\footnote{We borrow
the \emph{outlives} relation from~\cite{MIT03}. A comparison of our
approach with~\cite{MIT03} can be found in
\S~\ref{sec:related-work}.}, and any references from the stack region
to the transferable region are therefore safe. 
% In general, we can allow references from a region
% $R_0$ to another region $R_1$, provided that we establish that $R_1$ outlives
% $R_0$. 
Next, we note that by controlling the ``outlives'' relationships
between various regions, we only allow safe cross-region references,
while prohibiting unsafe ones. In the above example, an outlives
relationship \emph{from} the stack region \emph{to} the transferable
region means that the references in that direction are allowed, but
not the references in the opposite direction. In contrast, if an
\C{open} block of a transferable region $R_0$ is nested inside an
\C{open} block of another transferable region $R_1$, we do not
establish any outlives relationships, thus declaring our intention to
not allow any cross-region references between $R_0$ and $R_1$.
Finally, we observe that outlives relationships are established based
on the lexical structure of the program, hence a static type system
can enforce them effectively. By assigning region types to objects,
which capture the regions such objects are allocated in, and by
maintaining outlives relationships between various regions, we can
statically decide the safety of all references in the program.

Indeed, the utility of our approach described above is predicated
on the assumption that we can enforce certain invariants on
transferable regions. Firstly, a transferable region cannot be
transferred/freed inside an open block of that region (i.e, while it
is still open).  Secondly, a transferred/freed region cannot be
opened. These are typestate invariants on the transferable region
objects, which are hard to enforce statically due to the presence of
unrestricted aliasing. Techniques like linear types and unique
pointers can be used to restrict aliasing, but the constraints they
impose are often hard to program around. We therefore enforce
typestate invariants at runtime via lightweight checks. In
particular, we define an acceptable state transition discipline for
transferable regions (Fig.~\ref{fig:region-fsm}), and check, at
runtime, whether a given transition of a transferable region (\eg,
from \emph{open} state to \emph{freed} state) is valid or not. The
check is lightweight since it only involves checking a single tag that
captures the current state.  We believe that this is a reasonable
choice since regions are coarse-grained objects manipulated
infrequently, when compared to the fine-grained objects that are
present inside these regions, for which safety is enforced statically.
An added advantage of delegating the enforcement of typestate
invariants to runtime is that our region type system is
simple\footnote{Supplement lists all the type rules of our language,
which occupy less than a page.}, which made it possible to formulate a
type inference that completely eliminates the need to write region
type annotations. This, we believe, significantly reduces the
impediment to adopt our approach in practical setting.

@@
